Multivariate Density Estimation: Comparing Transformation Random Forest, Normalizing Flows and Copulas

### 2025-09-29 report/chapter02.Rnw
.

Transformation Random Forests \citep{hothorn2017transformation,hothorn2018conditional,hothorn2021transformation} fit into the same transport frame through the probability integral transform. Let $\widehat F_k(\cdot \mid u_{1:k-1})$ denote the strictly increasing conditional CDF returned by a TRTF for coordinate $k$. The induced triangular component is
\begin{equation}
  S_k(u_{1:k}) \;=\; \Phi^{-1}\!\Big(\widehat F_k(u_k \mid u_{1:k-1})\Big),\label{eq:transport-trtf-map}
\end{equation}
which maps conditionals to standard normal margins. In plain language, TRTF predicts a conditional CDF, then the probit transform places the result on the Gaussian reference scale. Differentiating $\Phi\!\big(S_k(u_{1:k})\big)=\widehat F_k(u_k \mid u_{1:k-1})$ with respect to $u_k$ yields
\begin{equation}
  \widehat \pi_k(u_k \mid u_{1:k-1}) \;=\; \varphi\!\big(S_k(u_{1:k})\big)\,\partial_{u_k}S_k(u_{1:k}),\label{eq:transport-trtf-likelihood}
\end{equation}
which is exactly the pullback factor in Equation~\eqref{eq:transport-likelihood}. Summing over $k$ recovers Equation~\eqref{eq:transport-likelihood} in standardized coordinates.

The additive-predictor TRTF used in this thesis yields a separable transport. Under the model
\begin{equation}
  \widehat F_k(u_k \mid u_{1:k-1}) \;=\; \Phi\!\big(h_k(u_k) + g_k(u_{1:k-1})\big),\label{eq:transport-trtf-additive}
\end{equation}
we obtain
\begin{equation}
  S_k(u_{1:k}) \;=\; h_k(u_k) + g_k(u_{1:k-1}),\qquad \partial_{u_k}S_k(u_{1:k}) \;=\; h_k'(u_k),\label{eq:transport-trtf-separable}
\end{equation}
so TRTF implements the same separable triangular likelihood as the direct parameterization in Equation~\eqref{eq:transport-separable}. Monotonicity in $u_k$ holds by construction, the Jacobian depends only on $u_k$, and inversion proceeds by back-substitution identical to the separable map. This equivalence underpins the empirical comparisons in Chapter~\ref{ch:dataanalysis}.

@@

A low-dimensional nonparametric variant avoids elliptical assumptions at small $K$. We fit a kernel density $\widehat f_Z$ on $z=\Phi^{-1}(v)$ and recover the copula density by
\begin{equation}
  c(v) \;=\; \frac{\widehat f_Z\!\big(\Phi^{-1}(v)\big)}{\prod_{k=1}^{K} \varphi\!\big(\Phi^{-1}(v_k)\big)},\label{eq:copula-kde}
\end{equation}
which applies the change of variables from $z$ back to $v$ and yields a proper copula density. In words, the kernel density models the joint shape of the probit scores, and division by the product of standard normal densities restores the unit-cube scale. This approach is viable only for small $K$, where kernel density estimation remains accurate and stable. Chapter~\ref{ch:dataanalysis} employs it strictly as a diagnostic baseline.

>>>>

Transformation Random Forests \citep{hothorn2017transformation,hothorn2018conditional,hothorn2021transformation} fit into the same transport frame through the probability integral transform. Let $\widehat F_k(\cdot \mid u_{1:k-1})$ denote the strictly increasing conditional CDF returned by a TRTF for coordinate $k$. (In practice, forest CDFs can be stepwise; we assume a measurable, strictly increasing version after standard monotone smoothing so that inversion and derivatives are well-defined.) The induced triangular component is
\begin{equation}
  S_k(u_{1:k}) \;=\; \Phi^{-1}\!\Big(\widehat F_k(u_k \mid u_{1:k-1})\Big),\label{eq:transport-trtf-map}
\end{equation}
which maps conditionals to standard normal margins. In plain language, TRTF predicts a conditional CDF, then the probit transform places the result on the Gaussian reference scale. Differentiating $\Phi\!\big(S_k(u_{1:k})\big)=\widehat F_k(u_k \mid u_{1:k-1})$ with respect to $u_k$ yields
\begin{equation}
  \widehat \pi_k(u_k \mid u_{1:k-1}) \;=\; \varphi\!\big(S_k(u_{1:k})\big)\,\partial_{u_k}S_k(u_{1:k}),\label{eq:transport-trtf-likelihood}
\end{equation}
which is exactly the pullback factor in Equation~\eqref{eq:transport-likelihood}. Summing over $k$ recovers Equation~\eqref{eq:transport-likelihood} in standardized coordinates.

The additive-predictor TRTF used in this thesis yields a separable transport. Under the model
\begin{equation}
  \widehat F_k(u_k \mid u_{1:k-1}) \;=\; \Phi\!\big(h_k(u_k) + g_k(u_{1:k-1})\big),\label{eq:transport-trtf-additive}
\end{equation}
we obtain
\begin{equation}
  S_k(u_{1:k}) \;=\; h_k(u_k) + g_k(u_{1:k-1}),\qquad \partial_{u_k}S_k(u_{1:k}) \;=\; h_k'(u_k),\label{eq:transport-trtf-separable}
\end{equation}
so TRTF implements the same separable triangular likelihood as the direct parameterization in Equation~\eqref{eq:transport-separable}. The map is monotone in $u_k$ by construction, the Jacobian depends only on $u_k$, and inversion proceeds by back-substitution identical to the separable map. This equivalence underpins the empirical comparisons in Chapter~\ref{ch:dataanalysis}.

@@

A low-dimensional nonparametric variant avoids elliptical assumptions at small $K$. We fit a kernel density $\widehat f_Z$ on $z=\Phi^{-1}(v)$ and recover the copula density by
\begin{equation}
  c(v) \;=\; \frac{\widehat f_Z\!\big(\Phi^{-1}(v)\big)}{\prod_{k=1}^{K} \varphi\!\big(\Phi^{-1}(v_k)\big)},\label{eq:copula-kde}
\end{equation}
which applies the change of variables from $z$ back to $v$ and yields a proper copula density. In words, the kernel density models the joint shape of the probit scores, and division by the product of standard normal densities restores the unit-cube scale. This approach is viable only for small $K$, where kernel density estimation remains accurate and stable. We implement this baseline via the \texttt{kdecopula} package \citep{nagler2017kdecopula}. Chapter~\ref{ch:dataanalysis} employs it strictly as a diagnostic baseline.
